### 2019-06-24 submission/manuscript.Rmd
Fix paragraph spacing and make some edits

## Importance
Antibiotics are a major risk factor for *Clostridium difficile* infections (CDIs), but it is less clear what other factors contribute to CDI incidence, severity, and recurrence. Interestingly, other medications besides antibiotics have been shown to alter the microbiota. Proton pump inhibitors (PPIs) were associated with alterations in the human intestinal microbiota through observational and interventional stduies and PPI use has also been associated with CDIs. We evaluated the ability of the PPI omeprazole to alter the murine intestinal microbiota and disrupt resistance to *C. difficile*. We found PPI treatment had minimal impact on the murine intestinal microbiota and was not enough to promote *C. difficile* colonization after challenge. Further studies are needed to determine whether other factors such as the composition of the starting bacterial community, comorbidities, and use of additional medications contribute to the association between PPIs and CDIs.

@@

## Introduction
Proton pump inhibitors (PPIs) are one of the most widely used drugs and are available over the counter [@Singh2018]. PPIs are approved to treat a variety of conditions including gastroesophageal reflux disease, ulcers, or as prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated gastrointestinal bleeding [@nehra2018proton]. Unfortunately, there have been a number of adverse events associated with their use including small intestinal bacterial overgrowth, decreased calcium absorption, and increased risk of enteric infections [@nehra2018proton]. There’s also been a number of large cohort studies as well as interventional clinical trials that have demonstrated specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018]. PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018]. Human fecal microbiota changes with PPI use include increases in Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae with decreases in Ruminococcaceae [@Imhann2017]. An *in vitro screen*, demonstrated that 3 PPIs inhibit the growth of 7-8 strains of Lachnospiraceae and Ruminococcaceae, which represented ~16% of the 40 intestinal bacteria strains that were screened [@maier2018extensive].
Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridium difficile* infections (CDIs) [@Schubert2015]. It is less clear whether other human medications that impact the microbiota also influence *C. difficile* colonization resistance. Multiple epidemiological studies have suggested an association between PPI use and incidence or recurrence of CDIs [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019]. However, most of the studies suggesting a link between PPIs and *C. difficile* are retrospective and did not evaluate the microbiome [@tariq2017association; @nehra2018proton; @Elias2019]. Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use play a role in the association between PPIs and CDI incidence or recurrence. Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDI infections or recurrent CDIs [@Elias2019; @tariq2017association]. 
Although there have been a few *C. difficile* mouse model studies that have demonstrated PPIs have some effect on CDIs with or without additional antibiotic treatment, none of these studies looked at the interplay with the intestinal microbiome [@hung2015proton; @Kaur2007; @kaur2011effect]. Here, we evaluated the impact of a PPI on the murine microbiome and susceptibility to *C. difficile* colonization in relation to clindamycin, an antibiotic that perturbs the microbiome enough to allow *C. difficile* to colonize but is mild enough that *C. difficile* is cleared within 10 days [@Jenior2018].

@@

**Murine fecal microbiomes were minimally affected by PPI treatment **
To test if PPI treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice daily with a high dose of omeprazole for 7 days before *C. difficile* challenge (Figure 1A) and compared the PPI-treated mice to a group that received the antibiotic, clindamycin, and a group that received both PPI and clindamycin treatment. A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of PPI-treated mice remained relatively unchanged (Figure 1B). Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes that have previously been impacted by PPIs in human studies, but we see no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 1C).In contrast, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the untreated and PPI-treated mice the day after treatment (Figure 1D).

**PPI-treatment did not promote susceptibility to *C. difficile infection* in mice**
After 7 days of PPI treatment or 1 day after clindamycin treatment, mice were challenged with *C. difficile* 630 spores. While all 4 of the clindamycin-treated mice were colonized with *C. difficile*, all of the PPI-treated mice were resistant to *C. difficile* colonization (Figure 2A). Interestingly, only 1 cage of the PPI and clindamycin-treated mice were colonized, while the other cage was not (Figure 2A). PCoAs of the bacterial communities from the 3 groups of mice after *C. difficile* challenge, showed that the greatest shifts in bacterial communities occurred in the clindamycin-treated mice (Figure 2B). Within 5 days, all of the mice had cleared *C. difficile*, suggesting there was no difference in rate of clearance for the mice that were initially colonized with *C. difficile*. Our results suggest that PPI treatment alone had minor effects on microbial community resistance to *C. difficile* colonization in mice. Instead most of the differences between our 3 treatment groups appear to be driven by clindamycin administration and included decreased *Alistipes*, *Barnesiella*, Porhyromonadaceae, Ruminococcaceae [@Schubert2015], taxa previously found to be altered in clindamycin-treated mice that were challenged with *C. difficile*.

@@

Our findings that PPI treatment had minimal impact on the fecal microbiome are comparable to several other mouse studies that indicate PPIs have more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018]. We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies, but a closer examination suggests that the significance of these associations is driven by some individuals more than others and the effect size can be relatively small with overall differences on a PCoA plot difficult to distinguish [@Freedberg2015, @Imhann2015]. Also, most of the studies to date compare different individuals (PPI users versus non-users) or the same individuals before and after treatment [@Naito2018; @Imhann2017], unfortunately these limited microbiota snapshots may be skewed by day-to-day microbiota variations [@LlornsRico2019], which become apparent in the longitudinally sampled mice. 
Previous work that administered the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C. difficile* demonstrated that PPI treatment alone resulted in detectable *C. difficile* in the stool 1 week after challenge, but also showed there was detectable *C. difficile* in mice not treated with antibiotics [@Kaur2007; @kaur2011effect]. The presence of *C. difficile* in mice that were not treated with either antibiotics or PPIs could be attributed to the higher dose of vegetative cells (10^8^ CFU) and the different *C. difficile* strain used to challenge the mice and may partly explain their observation of *C. difficile* in PPI-treated mice [@Kaur2007; @kaur2011effect]. We have previously shown mice from are colony that are not given antibiotics are resistant to *C. difficile* 630 when challenged with 10^3^ spores [@Jenior2017]. The other group also used a higher dose of 10^6^ spores or 10^7^ vegetative cells to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton]. Our study extended previous work examining PPIs and *C. difficile* in mice to examine the contribution of the intestinal microbiota. We found the PPI omeprazole had minimal impact on the murine intestinal microbiota and in contrast to previous work, PPIs did not promote *C. difficile* colonization. Our 16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare taxa in our C57BL/6 mouse colony. If these taxa are important contributors to the associations between PPIs and CDIs in humans, this may be part of the reason we did not see an effect in our CDI mouse model. There could also be differences at the species level for taxa that are altered with PPIs in humans compared to mice, which may also contribute to the association. Other murine starting communities or the addition of human PPI treatment-associated strains may be needed to determine whether PPIs can alter taxa and subsequently promote CDIs.  

>>>>

## Importance
Antibiotics are a major risk factor for *Clostridium difficile* infections (CDIs), but it is less clear what other factors contribute to CDI incidence and recurrence. Interestingly, other medications besides antibiotics have been shown to alter the microbiota. Proton pump inhibitors (PPIs) are associated with alterations in the human intestinal microbiota through observational and interventional stduies and PPI use has also been associated with CDIs. We evaluated the ability of the PPI omeprazole to alter the murine intestinal microbiota and disrupt colonization resistance to *C. difficile*. We found PPI treatment had minimal impact on the murine intestinal microbiota and was not enough to promote *C. difficile* colonization after challenge. Further studies are needed to determine whether other factors such as the composition of the starting bacterial community, comorbidities, and use of additional medications contribute to the association between PPIs and CDIs.

@@

## Introduction
Proton pump inhibitors (PPIs) are one of the most widely used drugs and are available over the counter [@Singh2018]. PPIs are approved to treat a variety of conditions including gastroesophageal reflux disease, ulcers, or as prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated gastrointestinal bleeding [@nehra2018proton]. Unfortunately, there have been a number of adverse events associated with their use including small intestinal bacterial overgrowth, decreased calcium absorption, and increased risk of enteric infections [@nehra2018proton]. There’s also been a number of large cohort studies as well as interventional clinical trials that have demonstrated specific alterations in the intestinal microbiome were associated with PPI use [@Imhann2017; @Naito2018]. PPI-associated microbiota changes have been attributed to the ability of PPIs to increase stomach acid pH which may promote the survival of oral and pathogenic bacteria [@Imhann2017; @Naito2018]. Human fecal microbiota changes with PPI use include increases in Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae with decreases in Ruminococcaceae [@Imhann2017]. An *in vitro screen*, demonstrated that 3 PPIs inhibit the growth of 7-8 strains of Lachnospiraceae and Ruminococcaceae, which represented ~16% of the 40 intestinal bacteria strains that were screened [@maier2018extensive].  
Antibiotics have a large impact on the intestinal microbiome and are a primary risk factor for developing *Clostridium difficile* infections (CDIs) [@Schubert2015]. It is less clear whether other human medications that impact the microbiota also influence *C. difficile* colonization resistance. Multiple epidemiological studies have suggested an association between PPI use and incidence or recurrence of CDIs [@tariq2017association; @nehra2018proton; @Naito2018; @Elias2019]. However, most of the studies suggesting a link between PPIs and *C. difficile* are retrospective and did not evaluate the microbiome [@tariq2017association; @nehra2018proton; @Elias2019]. Thus, it is unclear whether the gastrointestinal microbiome changes associated with PPI use play a role in the association between PPIs and CDI incidence or recurrence. Additionally, epidemiological studies have a limited capacity to address potential confounders and comorbidities in patients that were on PPIs and developed CDIs or recurrent CDIs [@Elias2019; @tariq2017association].    
Although there have been a few *C. difficile* mouse model studies that have demonstrated PPIs have some effect on CDIs with or without additional antibiotic treatment, none of these studies looked at the interplay with the intestinal microbiome [@hung2015proton; @Kaur2007; @kaur2011effect]. Here, we evaluated the impact of a PPI on the murine microbiome and susceptibility to *C. difficile* colonization in relation to clindamycin, an antibiotic that perturbs the microbiome enough to allow *C. difficile* to colonize but is mild enough that *C. difficile* is cleared within 10 days [@Jenior2018].

@@

**Murine fecal microbiomes were minimally affected by PPI treatment **
To test if PPI treatment alters the microbiome and promotes susceptibility to CDIs, we gavaged mice daily with a high dose of omeprazole for 7 days before *C. difficile* challenge (Figure 1A) and compared the PPI-treated mice to a group that received the antibiotic, clindamycin, and a group that received both PPI and clindamycin treatment. A principle coordinates analysis (PCoA) of the Bray-Curtis distances over the initial 7 days of treatment revealed the bacterial communities of PPI-treated mice remained relatively unchanged (Figure 1B). Enterococcaceae, Lactobacillaceae, Micrococcaceae, Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes that have previously been impacted by PPIs in human studies, but we see no significant fluctuations in the stool of our PPI-treated mice throughout the course of the 16-day experiment (Figure 1C). In contrast, the microbiomes from the 2 groups of mice that received clindamycin started to shift away from the untreated and PPI-treated mice the day after treatment (Figure 1D).

**PPI-treatment did not promote susceptibility to *C. difficile infection* in mice**
After 7 days of PPI treatment or 1 day after clindamycin treatment, mice were challenged with *C. difficile* 630 spores. While all 4 of the clindamycin-treated mice were colonized with *C. difficile*, all of the PPI-treated mice were resistant to *C. difficile* colonization (Figure 2A). Interestingly, only 1 cage of the PPI and clindamycin-treated mice were colonized, while the other cage was not (Figure 2A). PCoAs of the bacterial communities from the 3 groups of mice after *C. difficile* challenge, showed that the greatest shifts in bacterial communities occurred in the clindamycin-treated mice (Figure 2B). Within 5 days, all of the mice had cleared *C. difficile*, suggesting there was no difference in rate of clearance for the mice that were initially colonized with *C. difficile* despite continuing to receive PPI treatment throughout the course of the experiment. Our results suggest that PPI treatment alone had minor effects on microbial community resistance to *C. difficile* colonization in mice. Instead most of the differences between our 3 treatment groups appear to be driven by clindamycin administration and included decreased *Alistipes*, *Barnesiella*, Porhyromonadaceae, Ruminococcaceae, taxa previously found to be altered in clindamycin-treated mice that were challenged with *C. difficile* [@Schubert2015].

@@

Our findings that PPI treatment had minimal impact on the fecal microbiome are comparable to another PPI mouse study that indicated PPIs have more of an effect on the small intestinal microbiota compared to the fecal microbiota [@Yasutomi2018]. We did not find significant changes for the taxa observed to be significantly impacted by PPI use in human studies, but a closer examination suggests that the significance of these associations in humans is driven by some individuals more than others and the effect size can be relatively small with overall differences on a PCoA plot difficult to distinguish [@Freedberg2015, @Imhann2015]. Also, most of the studies to date compare different individuals (PPI users versus non-users) or the same individuals before and after treatment [@Naito2018; @Imhann2017], unfortunately these limited microbiota snapshots may be skewed by day-to-day microbiota variations [@LlornsRico2019] that would be revealed with more frequent longitudinal sampling.   
Previous work that administered the PPI lansoprazole daily for 2 weeks to mice and then challenged with *C. difficile* demonstrated that PPI treatment alone resulted in detectable *C. difficile* in the stool 1 week after challenge, but also showed there was detectable *C. difficile* in mice not treated with antibiotics [@Kaur2007; @kaur2011effect]. The presence of *C. difficile* in mice that were not treated with either antibiotics or PPIs could be attributed to the higher dose of vegetative cells (10^8^ CFU) and the different *C. difficile* strain used to challenge the mice and may partly explain their observation of *C. difficile* in PPI-treated mice [@Kaur2007; @kaur2011effect]. We have previously shown mice from are colony that are not given antibiotics are resistant to *C. difficile* 630 when challenged with 10^3^ spores [@Jenior2017]. Another group also used a higher dose of 10^6^ spores or 10^7^ vegetative cells and a different *C. difficile* strain to challenge antibiotic-treated mice or mice treated with both the PPI esomeprazole for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated mice developed more severe CDIs [@hung2015proton].  
Our study extended previous work examining PPIs and *C. difficile* in mice to examine the contribution of the intestinal microbiota. We found the PPI omeprazole had minimal impact on the murine intestinal microbiota and in contrast to previous work, PPIs did not promote *C. difficile* colonization. Our 16S rRNA sequencing suggested that *Streptococcus* and *Enterococcus* are rare taxa in our C57BL/6 mouse colony. If these taxa are important contributors to the associations between PPIs and CDIs in humans, this may be part of the reason we did not see an effect in our CDI mouse model. There could also be differences at the species level for taxa that are altered with PPIs in humans compared to mice, which may also contribute to the association. Other murine starting communities or the addition of human PPI treatment-associated strains may be needed to determine whether PPIs can alter taxa and subsequently promote CDIs.  

### 2019-06-24 submission/manuscript.tex
Fix paragraph spacing and make some edits

Antibiotics are a major risk factor for \emph{Clostridium difficile}
infections (CDIs), but it is less clear what other factors contribute to
CDI incidence, severity, and recurrence. Interestingly, other
medications besides antibiotics have been shown to alter the microbiota.
Proton pump inhibitors (PPIs) were associated with alterations in the
human intestinal microbiota through observational and interventional
stduies and PPI use has also been associated with CDIs. We evaluated the
ability of the PPI omeprazole to alter the murine intestinal microbiota
and disrupt resistance to \emph{C. difficile}. We found PPI treatment
had minimal impact on the murine intestinal microbiota and was not
enough to promote \emph{C. difficile} colonization after challenge.
Further studies are needed to determine whether other factors such as
the composition of the starting bacterial community, comorbidities, and
use of additional medications contribute to the association between PPIs
and CDIs.

@@

Proton pump inhibitors (PPIs) are one of the most widely used drugs and
are available over the counter (1). PPIs are approved to treat a variety
of conditions including gastroesophageal reflux disease, ulcers, or as
prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated
gastrointestinal bleeding (2). Unfortunately, there have been a number
of adverse events associated with their use including small intestinal
bacterial overgrowth, decreased calcium absorption, and increased risk
of enteric infections (2). There's also been a number of large cohort
studies as well as interventional clinical trials that have demonstrated
specific alterations in the intestinal microbiome were associated with
PPI use (3, 4). PPI-associated microbiota changes have been attributed
to the ability of PPIs to increase stomach acid pH which may promote the
survival of oral and pathogenic bacteria (3, 4). Human fecal microbiota
changes with PPI use include increases in Enterococcaceae,
Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae
with decreases in Ruminococcaceae (3). An \emph{in vitro screen},
demonstrated that 3 PPIs inhibit the growth of 7-8 strains of
Lachnospiraceae and Ruminococcaceae, which represented
\textasciitilde{}16\% of the 40 intestinal bacteria strains that were
screened (5). Antibiotics have a large impact on the intestinal
microbiome and are a primary risk factor for developing
\emph{Clostridium difficile} infections (CDIs) (6). It is less clear
whether other human medications that impact the microbiota also
influence \emph{C. difficile} colonization resistance. Multiple
epidemiological studies have suggested an association between PPI use
and incidence or recurrence of CDIs (2, 4, 7, 8). However, most of the
studies suggesting a link between PPIs and \emph{C. difficile} are
retrospective and did not evaluate the microbiome (2, 7, 8). Thus, it is
unclear whether the gastrointestinal microbiome changes associated with
PPI use play a role in the association between PPIs and CDI incidence or
recurrence. Additionally, epidemiological studies have a limited
capacity to address potential confounders and comorbidities in patients
that were on PPIs and developed CDI infections or recurrent CDIs (7, 8).
Although there have been a few \emph{C. difficile} mouse model studies
that have demonstrated PPIs have some effect on CDIs with or without
additional antibiotic treatment, none of these studies looked at the
interplay with the intestinal microbiome (9--11). Here, we evaluated the
impact of a PPI on the murine microbiome and susceptibility to \emph{C.
difficile} colonization in relation to clindamycin, an antibiotic that
perturbs the microbiome enough to allow \emph{C. difficile} to colonize
but is mild enough that \emph{C. difficile} is cleared within 10 days
(12).

@@

\textbf{Murine fecal microbiomes were minimally affected by PPI
treatment } To test if PPI treatment alters the microbiome and promotes
susceptibility to CDIs, we gavaged mice daily with a high dose of
omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A)
and compared the PPI-treated mice to a group that received the
antibiotic, clindamycin, and a group that received both PPI and
clindamycin treatment. A principle coordinates analysis (PCoA) of the
Bray-Curtis distances over the initial 7 days of treatment revealed the
bacterial communities of PPI-treated mice remained relatively unchanged
(Figure 1B). Enterococcaceae, Lactobacillaceae, Micrococcaceae,
Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes
that have previously been impacted by PPIs in human studies, but we see
no significant fluctuations in the stool of our PPI-treated mice
throughout the course of the 16-day experiment (Figure 1C).In contrast,
the microbiomes from the 2 groups of mice that received clindamycin
started to shift away from the untreated and PPI-treated mice the day
after treatment (Figure 1D).

\textbf{PPI-treatment did not promote susceptibility to \emph{C.
difficile infection} in mice} After 7 days of PPI treatment or 1 day
after clindamycin treatment, mice were challenged with \emph{C.
difficile} 630 spores. While all 4 of the clindamycin-treated mice were
colonized with \emph{C. difficile}, all of the PPI-treated mice were
resistant to \emph{C. difficile} colonization (Figure 2A).
Interestingly, only 1 cage of the PPI and clindamycin-treated mice were
colonized, while the other cage was not (Figure 2A). PCoAs of the
bacterial communities from the 3 groups of mice after \emph{C.
difficile} challenge, showed that the greatest shifts in bacterial
communities occurred in the clindamycin-treated mice (Figure 2B). Within
5 days, all of the mice had cleared \emph{C. difficile}, suggesting
there was no difference in rate of clearance for the mice that were
initially colonized with \emph{C. difficile}. Our results suggest that
PPI treatment alone had minor effects on microbial community resistance
to \emph{C. difficile} colonization in mice. Instead most of the
differences between our 3 treatment groups appear to be driven by
clindamycin administration and included decreased \emph{Alistipes},
\emph{Barnesiella}, Porhyromonadaceae, Ruminococcaceae (6), taxa
previously found to be altered in clindamycin-treated mice that were
challenged with \emph{C. difficile}.

@@

with \emph{C. difficile} while the other cage remained resistant. Our
findings that PPI treatment had minimal impact on the fecal microbiome
are comparable to several other mouse studies that indicate PPIs have
more of an effect on the small intestinal microbiota compared to the
fecal microbiota (13). We did not find significant changes for the taxa
observed to be significantly impacted by PPI use in human studies, but a
closer examination suggests that the significance of these associations
is driven by some individuals more than others and the effect size can
be relatively small with overall differences on a PCoA plot difficult to
distinguish (14). Also, most of the studies to date compare different
individuals (PPI users versus non-users) or the same individuals before
and after treatment (3, 4), unfortunately these limited microbiota
snapshots may be skewed by day-to-day microbiota variations (16), which
become apparent in the longitudinally sampled mice. Previous work that
administered the PPI lansoprazole daily for 2 weeks to mice and then
challenged with \emph{C. difficile} demonstrated that PPI treatment
alone resulted in detectable \emph{C. difficile} in the stool 1 week
after challenge, but also showed there was detectable \emph{C.
difficile} in mice not treated with antibiotics (10, 11). The presence
of \emph{C. difficile} in mice that were not treated with either
antibiotics or PPIs could be attributed to the higher dose of vegetative
cells (10\textsuperscript{8} CFU) and the different \emph{C. difficile}
strain used to challenge the mice and may partly explain their
observation of \emph{C. difficile} in PPI-treated mice (10, 11). We have
previously shown mice from are colony that are not given antibiotics are
resistant to \emph{C. difficile} 630 when challenged with
10\textsuperscript{3} spores (17). The other group also used a higher

@@

cells to challenge antibiotic-treated mice or mice treated with both the
PPI esomeprazole for 2 days and antibiotics and demonstrated the
antibiotic/PPI-treated mice developed more severe CDIs (9). Our study
extended previous work examining PPIs and \emph{C. difficile} in mice to
examine the contribution of the intestinal microbiota. We found the PPI
omeprazole had minimal impact on the murine intestinal microbiota and in
contrast to previous work, PPIs did not promote \emph{C. difficile}
colonization. Our 16S rRNA sequencing suggested that

>>>>

Antibiotics are a major risk factor for \emph{Clostridium difficile}
infections (CDIs), but it is less clear what other factors contribute to
CDI incidence and recurrence. Interestingly, other medications besides
antibiotics have been shown to alter the microbiota. Proton pump
inhibitors (PPIs) are associated with alterations in the human
intestinal microbiota through observational and interventional stduies
and PPI use has also been associated with CDIs. We evaluated the ability
of the PPI omeprazole to alter the murine intestinal microbiota and
disrupt colonization resistance to \emph{C. difficile}. We found PPI
treatment had minimal impact on the murine intestinal microbiota and was
not enough to promote \emph{C. difficile} colonization after challenge.
Further studies are needed to determine whether other factors such as
the composition of the starting bacterial community, comorbidities, and
use of additional medications contribute to the association between PPIs
and CDIs.

@@

Proton pump inhibitors (PPIs) are one of the most widely used drugs and
are available over the counter (1). PPIs are approved to treat a variety
of conditions including gastroesophageal reflux disease, ulcers, or as
prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated
gastrointestinal bleeding (2). Unfortunately, there have been a number
of adverse events associated with their use including small intestinal
bacterial overgrowth, decreased calcium absorption, and increased risk
of enteric infections (2). There's also been a number of large cohort
studies as well as interventional clinical trials that have demonstrated
specific alterations in the intestinal microbiome were associated with
PPI use (3, 4). PPI-associated microbiota changes have been attributed
to the ability of PPIs to increase stomach acid pH which may promote the
survival of oral and pathogenic bacteria (3, 4). Human fecal microbiota
changes with PPI use include increases in Enterococcaceae,
Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae
with decreases in Ruminococcaceae (3). An \emph{in vitro screen},
demonstrated that 3 PPIs inhibit the growth of 7-8 strains of
Lachnospiraceae and Ruminococcaceae, which represented
\textasciitilde{}16\% of the 40 intestinal bacteria strains that were
screened (5).\\
Antibiotics have a large impact on the intestinal microbiome and are a
primary risk factor for developing \emph{Clostridium difficile}
infections (CDIs) (6). It is less clear whether other human medications
that impact the microbiota also influence \emph{C. difficile}
colonization resistance. Multiple epidemiological studies have suggested
an association between PPI use and incidence or recurrence of CDIs (2,
4, 7, 8). However, most of the studies suggesting a link between PPIs
and \emph{C. difficile} are retrospective and did not evaluate the
microbiome (2, 7, 8). Thus, it is unclear whether the gastrointestinal
microbiome changes associated with PPI use play a role in the
association between PPIs and CDI incidence or recurrence. Additionally,
epidemiological studies have a limited capacity to address potential
confounders and comorbidities in patients that were on PPIs and
developed CDIs or recurrent CDIs (7, 8).\\
Although there have been a few \emph{C. difficile} mouse model studies
that have demonstrated PPIs have some effect on CDIs with or without
additional antibiotic treatment, none of these studies looked at the
interplay with the intestinal microbiome (9--11). Here, we evaluated the
impact of a PPI on the murine microbiome and susceptibility to \emph{C.
difficile} colonization in relation to clindamycin, an antibiotic that
perturbs the microbiome enough to allow \emph{C. difficile} to colonize
but is mild enough that \emph{C. difficile} is cleared within 10 days
(12).

@@

\textbf{Murine fecal microbiomes were minimally affected by PPI
treatment } To test if PPI treatment alters the microbiome and promotes
susceptibility to CDIs, we gavaged mice daily with a high dose of
omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A)
and compared the PPI-treated mice to a group that received the
antibiotic, clindamycin, and a group that received both PPI and
clindamycin treatment. A principle coordinates analysis (PCoA) of the
Bray-Curtis distances over the initial 7 days of treatment revealed the
bacterial communities of PPI-treated mice remained relatively unchanged
(Figure 1B). Enterococcaceae, Lactobacillaceae, Micrococcaceae,
Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes
that have previously been impacted by PPIs in human studies, but we see
no significant fluctuations in the stool of our PPI-treated mice
throughout the course of the 16-day experiment (Figure 1C). In contrast,
the microbiomes from the 2 groups of mice that received clindamycin
started to shift away from the untreated and PPI-treated mice the day
after treatment (Figure 1D).

\textbf{PPI-treatment did not promote susceptibility to \emph{C.
difficile infection} in mice} After 7 days of PPI treatment or 1 day
after clindamycin treatment, mice were challenged with \emph{C.
difficile} 630 spores. While all 4 of the clindamycin-treated mice were
colonized with \emph{C. difficile}, all of the PPI-treated mice were
resistant to \emph{C. difficile} colonization (Figure 2A).
Interestingly, only 1 cage of the PPI and clindamycin-treated mice were
colonized, while the other cage was not (Figure 2A). PCoAs of the
bacterial communities from the 3 groups of mice after \emph{C.
difficile} challenge, showed that the greatest shifts in bacterial
communities occurred in the clindamycin-treated mice (Figure 2B). Within
5 days, all of the mice had cleared \emph{C. difficile}, suggesting
there was no difference in rate of clearance for the mice that were
initially colonized with \emph{C. difficile} despite continuing to
receive PPI treatment throughout the course of the experiment. Our
results suggest that PPI treatment alone had minor effects on microbial
community resistance to \emph{C. difficile} colonization in mice.
Instead most of the differences between our 3 treatment groups appear to
be driven by clindamycin administration and included decreased
\emph{Alistipes}, \emph{Barnesiella}, Porhyromonadaceae,
Ruminococcaceae, taxa previously found to be altered in
clindamycin-treated mice that were challenged with \emph{C. difficile}
(6).

@@

with \emph{C. difficile} while the other cage remained resistant.\\
Our findings that PPI treatment had minimal impact on the fecal
microbiome are comparable to another PPI mouse study that indicated PPIs
have more of an effect on the small intestinal microbiota compared to
the fecal microbiota (13). We did not find significant changes for the
taxa observed to be significantly impacted by PPI use in human studies,
but a closer examination suggests that the significance of these
associations in humans is driven by some individuals more than others
and the effect size can be relatively small with overall differences on
a PCoA plot difficult to distinguish (14). Also, most of the studies to
date compare different individuals (PPI users versus non-users) or the
same individuals before and after treatment (3, 4), unfortunately these
limited microbiota snapshots may be skewed by day-to-day microbiota
variations (16) that would be revealed with more frequent longitudinal
sampling.\\
Previous work that administered the PPI lansoprazole daily for 2 weeks
to mice and then challenged with \emph{C. difficile} demonstrated that
PPI treatment alone resulted in detectable \emph{C. difficile} in the
stool 1 week after challenge, but also showed there was detectable
\emph{C. difficile} in mice not treated with antibiotics (10, 11). The
presence of \emph{C. difficile} in mice that were not treated with
either antibiotics or PPIs could be attributed to the higher dose of
vegetative cells (10\textsuperscript{8} CFU) and the different \emph{C.
difficile} strain used to challenge the mice and may partly explain
their observation of \emph{C. difficile} in PPI-treated mice (10, 11).
We have previously shown mice from are colony that are not given
antibiotics are resistant to \emph{C. difficile} 630 when challenged
with 10\textsuperscript{3} spores (17). Another group also used a higher

@@

cells and a different \emph{C. difficile} strain to challenge
antibiotic-treated mice or mice treated with both the PPI esomeprazole
for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated
mice developed more severe CDIs (9).\\
Our study extended previous work examining PPIs and \emph{C. difficile}
in mice to examine the contribution of the intestinal microbiota. We
found the PPI omeprazole had minimal impact on the murine intestinal
microbiota and in contrast to previous work, PPIs did not promote
\emph{C. difficile} colonization. Our 16S rRNA sequencing suggested that
